---
title: "time_series_project"
author: "Elizabeth Binkina"
date: "3/28/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(tseries)
library(forecast)


df <- read.csv("drug_overdose.csv")
df

View(df)
```

```{r}
df_clean <- df %>% 
  select(TIME_PERIOD, ESTIMATE) %>%
  group_by(TIME_PERIOD) %>%
  summarize(Avg_Estimate = mean(ESTIMATE, na.rm = TRUE)) %>%
  arrange(TIME_PERIOD)

ts_data <- ts(df_clean$Avg_Estimate, start = min(df_clean$TIME_PERIOD), frequency=1)

plot.ts(ts_data, col = "darkblue", lwd=2, 
        main = "Drug Overdose Time Series", ylab = "Estimate", xlab = "Year")
```
```{r}
library(TSA)
library(fpp2)
ts.plot(log(ts_data), main = "Xt")
```



```{r}
adf.test(log(ts_data))
```

```{r}
#first-order differencing

ts_diff1 <- diff(log(ts_data))
plot.ts(ts_diff1, col="darkred", lwd=2, 
        main = "First-Order Differenced Series", ylab = "Differenced Estimate", xlab = "Year")
```

```{r}
adf.test(ts_diff1)
```

```{r}
ts_diff2 <- diff(ts_diff1)

# Plot second differenced series
plot.ts(ts_diff2, col = "darkgreen", lwd = 2,
        main = "Second-Order Differenced Series", ylab = "2nd Diff Estimate", xlab = "Year")

```

```{r}
adf.test(ts_diff2)
```


```{r}
library(TSA)
library(fpp2)

```

```{r}
par(mfrow=c(1,2))
acf(ts_diff2)
pacf(ts_diff2)
```

```{r}
par(mfrow=c(1,2))
acf(ts_data)
pacf(ts_data)
```


ACF plot tails off
PACF plot cuts off after lag 1
ARMA(1,0)
AR(1) is the most appropriate model 


```{r}
library(forecast)
ar1_model <- Arima(ts_diff2, order = c(1, 0, 0))
summary(ar1_model)
```

```{r}
library(forecast)
auto_model <- auto.arima(ts_data)
summary(auto_model)
```

ARIMA(0,2,1) model

0 AR terms (no autoregressive structure),

2 differences (data required second-order differencing for stationarity),

1 MA term (moving average of order 1).

Model Summary
The MA(1) coefficient is significant (-0.5221 with SE 0.2106), implying that the model captures short-term shocks well.

The log-likelihood, AIC, BIC are all acceptable for a basic model.

Residual ACF1 ≈ 0.0042 → residuals are essentially uncorrelated → model fits well.

#Forcasting future overdose rates using this model

```{r}
library(forecast)

# Forecast next 5 years
forecast_values <- forecast(auto_model, h = 5)

# Plot forecast
autoplot(forecast_values) +
  ggtitle("Forecast of Drug Overdose Rates") +
  ylab("Estimated Rate") +
  xlab("Year")

```

```{r}
checkresiduals(auto_model)
```
Residual Diagnostics Interpretation
1. Top Plot: Residuals Over Time
Residuals appear randomly scattered around zero, with no clear pattern.

Some higher residuals post-2020, but overall randomness suggests the model captures the structure well.

2. Bottom Left: ACF of Residuals
All autocorrelations are within the blue confidence bounds → no significant autocorrelation left.

This confirms the residuals are white noise, which is a good sign.

3. Bottom Right: Histogram of Residuals
Slight skew, but residuals roughly follow a normal distribution.

The red density line follows the bars fairly well.

No extreme outliers or fat tails to worry about.

Residuals are approximately uncorrelated and normally distributed, meeting assumptions for ARIMA modeling.

```{r}
library(forecast)
forecast_vals <- forecast(auto_model, h = 5)
autoplot(forecast_vals) +
  ggtitle("5-Year Forecast of Drug Overdose Rates") +
  ylab("Estimated Rate") + xlab("Year")
```

```{r}
# Example: split series
train <- window(ts_data, end = c(2018))
test <- window(ts_data, start = c(2019))

# Fit model on training
model_cv <- Arima(train, order = c(0,2,1))

# Forecast and compare
fc <- forecast(model_cv, h = length(test))
accuracy(fc, test)

```


```{r}
naive_model <- naive(ts_data)
rw_model <- rwf(ts_data, drift = TRUE)
accuracy(naive_model)
accuracy(rw_model)
```

```{r}
write.csv(as.data.frame(forecast_vals), "forecast_output.csv")
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```