---
title: "bitcoin_project"
author: "Elizabeth Binkina"
date: "9/22/2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#!. Intro 

#2. Load Data

```{r}
df <- read.csv("_5acbed1925f148fe950aa47000a74f15_bitcoin.csv")
View(df)
```

```{r}
library(zoo)
library(tidyverse)
library(glmnet)
library(dplyr)
library(tidyr)
library(leaps)
library(forecast)
library(car)
library(splines)
set.seed(598)
```

```{r}
df$Date <- as.Date(df$Date)
str(df)
```
#3. Data Cleaning and Reconstruction

```{r cleaning }
# Replace zeros with NA 
df <- df %>% mutate(btc_market_price = ifelse(btc_market_price <= 0,
                                              NA, btc_market_price))

# Sort by Date 
df <- df %>% arrange(Date)
df$btc_market_price <- na.approx(df$btc_market_price, na.rm = FALSE)

df <- df %>% drop_na(btc_market_price)
# Stabilize variance: log transform
df <- df %>% mutate(log_price = log(btc_market_price))

cat("Rows after cleaning:", nrow(df), "\n")
summary(df$btc_market_price)
```


```{r}
p1 <- ggplot(df, aes(Date, btc_market_price)) + geom_line() + 
  labs(title="BTC Price (USD)", y="Price (USD)")
p2 <- ggplot(df, aes(Date, log_price)) + geom_line() + 
  labs(title="BTC Log-Price", y="Log(Price)")

p1; p2
```

#4. Exploratory Data Analysis 

Outliers
```{r}
iqr_price <- IQR(df$btc_market_price)
q1_price <- quantile(df$btc_market_price, 0.25)
q3_price <- quantile(df$btc_market_price, 0.75)
lower <- q1_price - 1.5 * iqr_price
upper <- q3_price + 1.5 * iqr_price
price_outliers_pct <- mean(df$btc_market_price < lower | 
                             df$btc_market_price > upper)
cat("Proportion of price points outside IQR bounds:", 
    scales::percent(price_outliers_pct), "\n")
```

#5. Feature Engineering 

```{r features}
horizon <- 7      # predict t + 7 days
n_lags <- 14      # construct lag1..lag14

df_feat <- df %>% arrange(Date)
for(i in 1:n_lags) df_feat[[paste0("log_price_lag", i)]] <- dplyr::lag(df_feat$log_price, i)

df_feat <- df_feat %>%
  mutate(
    ma7 = zoo::rollmean(log_price, 7, fill = NA, align = "right"),
    ma14 = zoo::rollmean(log_price, 14, fill = NA, align = "right"),
    vol7 = zoo::rollapply(log_price, width = 7, FUN = sd, fill = NA, align = "right")
  )

# target is log price at t + horizon
df_feat <- df_feat %>% mutate(target_log_price = dplyr::lead(log_price, horizon))
df_feat <- df_feat %>% drop_na()
cat("Rows after feature engineering:", nrow(df_feat), "\n")
```

```{r}
str(df_feat)
```


```{r split}
#Time-Ordered train / validation / test splot
N <- nrow(df_feat)
train_end <- floor(0.70 * N)
valid_end <- floor(0.85 * N)

train <- df_feat[1:train_end, ]
valid <- df_feat[(train_end + 1):valid_end, ]
test  <- df_feat[(valid_end + 1):N, ]

cat("Train rows:", nrow(train), 
    "Validation rows:", nrow(valid), "Test rows:", nrow(test), "\n")
```

```{r modelmatrix}
preds_train <- train %>% select(-Date, -btc_market_price, -log_price, -target_log_price)
preds_valid <- valid %>% select(-Date, -btc_market_price, -log_price, -target_log_price)
preds_test  <- test  %>% select(-Date, -btc_market_price, -log_price, -target_log_price)

x_train <- model.matrix(~ . , data = preds_train)[, -1]
y_train <- train$target_log_price

x_valid <- model.matrix(~ . , data = preds_valid)[, -1]
y_valid <- valid$target_log_price

x_test <- model.matrix(~ . , data = preds_test)[, -1]
y_test <- test$target_log_price

common_cols <- intersect(colnames(x_train), colnames(x_valid))
common_cols <- intersect(common_cols, colnames(x_test))

x_train <- x_train[, common_cols, drop = FALSE]
x_valid <- x_valid[, common_cols, drop = FALSE]
x_test  <- x_test[, common_cols, drop = FALSE]

rmse <- function(actual, pred) sqrt(mean((actual - pred)^2, na.rm = TRUE))
rmse_price <- function(actual_log, pred_log) sqrt(mean
                                                  ((exp(actual_log) - exp(pred_log))^2, na.rm = TRUE))
mape <- function(actual_log, pred_log) mean(
  abs((exp(actual_log) - exp(pred_log)) / (exp(actual_log) + 1e-8)), na.rm = TRUE) * 100
```

#8.(lag-horizon)

```{r naive}
naive_pred_valid <- valid[[paste0("log_price_lag", horizon)]]
naive_rmse_valid <- rmse(y_valid, naive_pred_valid)
naive_rmse_price_valid <- rmse_price(y_valid, naive_pred_valid)
naive_mape_valid <- mape(y_valid, naive_pred_valid)

naive_pred_test <- test[[paste0("log_price_lag", horizon)]]
naive_rmse_test <- rmse(y_test, naive_pred_test)
naive_rmse_price_test <- rmse_price(y_test, naive_pred_test)
naive_mape_test <- mape(y_test, naive_pred_test)

cat("validation RMSE (log):", round(naive_rmse_valid, 5),
    " RMSE (price):", round(naive_rmse_price_valid, 2)," MAPE (%):", 
    round(naive_mape_valid, 3), "\n")
```

Model Selection 

# 9. OLS (manual chosen predictors) — train on training set, validate on validation

```{r}
ols_formula <- as.formula(paste("target_log_price ~",
                               paste(c(paste0("log_price_lag", 1:5), "ma7", "ma14", "vol7"), 
                                     collapse = " + ")))

lm_model <- lm(ols_formula, data = train)
summary(lm_model)
pred_val_ols <- predict(lm_model, newdata = valid)
ols_rmse_val <- rmse(y_valid, pred_val_ols)
ols_rmse_price_val <- rmse_price(y_valid, pred_val_ols)
ols_mape_val <- mape(y_valid, pred_val_ols)
cat("OLS validation RMSE (log):", round(ols_rmse_val,5), "price-RMSE:", 
    round(ols_rmse_price_val,2), "MAPE:", round(ols_mape_val,3), "\n")

```


#Multicollinearity Check (VIF)
```{r vif-check}
vif_values <- vif(lm_model)
print(vif_values)
```


# residual Diagnostics for Best Model (Replace `best_model` with chosen one)
```{r residual-diagnostics}
resid_best <- resid(lm_model)
acf(resid_best, main = "ACF of residuals (Best Model)")
Box.test(resid_best, lag = 20, type = "Ljung-Box")
```
### WILL Not be using OLS model due to collinearity


# 10. Best-subset selection (leaps) 

```{r bestsubset}
vars_for_subset <- c(paste0("log_price_lag", 1:14), "ma7", "ma14", "vol7")
subset_formula <- as.formula(paste("target_log_price ~", 
                                   paste(vars_for_subset, collapse = " + ")))

regfit <- regsubsets(subset_formula, data = train, nvmax = 10, method = "exhaustive")
reg_sum <- summary(regfit)

predict_regsubsets <- function(object, newdata, id) {
  coefi <- coef(object, id = id)
  mat <- model.matrix(as.formula(paste("~",
                                       paste(vars_for_subset, collapse = " + "))), newdata)
  mat <- mat[, names(coefi)[names(coefi) != "(Intercept)"], drop = FALSE]
  preds <- mat %*% coefi[names(coefi) != "(Intercept)"] + coefi["(Intercept)"]
  as.numeric(preds)
}

nv <- min(10, length(reg_sum$which))
valid_results <- tibble(size = 1:nv, 
                        RMSE = NA_real_, RMSE_price = NA_real_, MAPE = NA_real_)
for(i in seq_len(nv)){
  preds_i <- predict_regsubsets(regfit, valid, id = i)
  valid_results$RMSE[i] <- rmse(y_valid, preds_i)
  valid_results$RMSE_price[i] <- rmse_price(y_valid, preds_i)
  valid_results$MAPE[i] <- mape(y_valid, preds_i)
}
valid_results <- valid_results %>% arrange(RMSE)
valid_results

best_size <- valid_results$size[which.min(valid_results$RMSE)]
cat("Best subset size by validation RMSE:", best_size, "\n")
best_coeffs <- coef(regfit, id = best_size)
best_coeffs
```


```{r}
selected_vars <- names(best_coeffs)[-1]
best_formula <- as.formula(paste("target_log_price ~", paste(selected_vars, collapse = " + ")))
best_model_trainval <- lm(best_formula, data = bind_rows(train, valid))

pred_best_test <- predict(best_model_trainval, newdata = test)
best_rmse_test <- rmse(y_test, pred_best_test)
best_rmse_price_test <- rmse_price(y_test, pred_best_test)
best_mape_test <- mape(y_test, pred_best_test)

best_rmse_val <- valid_results$RMSE[valid_results$size == best_size]
best_rmse_price_val <- valid_results$RMSE_price[valid_results$size == best_size]
best_mape_val <- valid_results$MAPE[valid_results$size == best_size]

cat("Best-subset (size=", best_size,
    ") test RMSE (log):", round(best_rmse_test,5),
    "price-RMSE:", round(best_rmse_price_test,2), "MAPE:",
    round(best_mape_test,3), "\n")
```

# 11. Lasso (glmnet) — tune lambda using the validation set (time-aware)

```{r lasso}
lambda_grid <- 10^seq(-4, 0, length.out = 100)
lasso_fit <- glmnet(x_train, y_train, alpha = 1, lambda = lambda_grid)

val_preds_mat <- predict(lasso_fit, newx = x_valid)
val_rmse_by_lambda <- apply(val_preds_mat, 2, function(p) rmse(y_valid, p))

best_lambda_lasso <- lambda_grid[which.min(val_rmse_by_lambda)]
cat("Best lambda (Lasso):", best_lambda_lasso, "\n")

# refit 
x_train_valid <- rbind(x_train, x_valid)
y_train_valid <- c(y_train, y_valid)
final_lasso <- glmnet(x_train_valid, y_train_valid, alpha = 1,
                      lambda = best_lambda_lasso)

# validation
pred_lasso_val <- predict(lasso_fit, 
                          newx = x_valid, s = best_lambda_lasso)
lasso_rmse_val <- rmse(y_valid, pred_lasso_val)
lasso_rmse_price_val <- rmse_price(y_valid, pred_lasso_val)
lasso_mape_val <- mape(y_valid, pred_lasso_val)

cat("Lasso (validation) - RMSE (log):", round(lasso_rmse_val, 5),
"RMSE (price):", round(lasso_rmse_price_val, 2),
"MAPE (%):", round(lasso_mape_val, 3), "\n")

# non-zero coefficients
coef_final <- as.matrix(coef(final_lasso))
nonzero <- rownames(coef_final)[coef_final != 0]
print(nonzero)
```


# 12. Ridge Regression

```{r}
ridge_fit <- glmnet(x_train, y_train, alpha = 0, lambda = lambda_grid)
val_preds_ridge <- predict(ridge_fit, newx = x_valid)
val_rmse_ridge <- apply(val_preds_ridge, 2, function(p) rmse(y_valid, p))


best_lambda_ridge <- lambda_grid[which.min(val_rmse_ridge)]
cat("Best lambda (Ridge):", best_lambda_ridge, "\n")


final_ridge <- glmnet(x_train_valid, 
                      y_train_valid, alpha = 0, lambda = best_lambda_ridge)
pred_ridge_val <- predict(ridge_fit, newx = x_valid, s = best_lambda_ridge)
ridge_rmse_val <- rmse(y_valid, pred_ridge_val)
ridge_rmse_price_val <- rmse_price(y_valid, pred_ridge_val)
ridge_mape_val <- mape(y_valid, pred_ridge_val)


cat("Ridge (validation) - RMSE (log):", round(ridge_rmse_val, 5),
"RMSE (price):", round(ridge_rmse_price_val, 2),
"MAPE (%):", round(ridge_mape_val, 3), "\n")
```


#13. Test Set Evaluation

```{r test-eval}
# Best-subset refit on train+valid
top_vars <- names(best_coeffs)[-1]
best_formula <- as.formula(paste("target_log_price ~", paste(top_vars, collapse = "+")))
model_best_trainval <- lm(best_formula, data = bind_rows(train, valid))
pred_best_test <- predict(model_best_trainval, newdata = test)

# Predictions
pred_lasso_test <- as.numeric(predict(final_lasso, newx = x_test, s = best_lambda_lasso))
pred_ridge_test <- as.numeric(predict(final_ridge, newx = x_test, s = best_lambda_ridge))
naive_pred_test <- test[[paste0("log_price_lag", horizon)]]

# Metrics
test_results <- tibble(
Model = c("Naive", "Best-Subset", "Ridge", "Lasso"),
RMSE_log = c(rmse(y_test, naive_pred_test),
rmse(y_test, pred_best_test),
rmse(y_test, pred_ridge_test),
rmse(y_test, pred_lasso_test)),
RMSE_price = c(rmse_price(y_test, naive_pred_test),
rmse_price(y_test, pred_best_test),
rmse_price(y_test, pred_ridge_test),
rmse_price(y_test, pred_lasso_test)),
MAPE = c(mape(y_test, naive_pred_test),
mape(y_test, pred_best_test),
mape(y_test, pred_ridge_test),
mape(y_test, pred_lasso_test))
) %>% mutate_if(is.numeric, round, 5)


test_results
```


# 14. Visual Comparison

```{r}
df_plot <- tibble(
date = test$Date,
actual = y_test,
best = pred_best_test,
lasso = pred_lasso_test,
ridge = pred_ridge_test
)

df_plot <- na.omit(df_plot)
plot(df_plot$date, exp(df_plot$actual), type = "l", lwd = 2, col = "black",
ylab = "BTC Price (USD)", xlab = "Date",
main = "Test set: Actual vs Predicted")
lines(df_plot$date, exp(df_plot$best), col = "blue", lwd = 1.5)
lines(df_plot$date, exp(df_plot$lasso), col = "red", lwd = 1.5)
lines(df_plot$date, exp(df_plot$ridge), col = "green", lwd = 1.5)
legend("topleft", legend = c("Actual", "Best-Subset", "Lasso", "Ridge"),
col = c("black", "blue", "red", "green"), lty = 1, bty = "n")

```


# 15. Decision: 

```{r decision}
best_model_row <- test_results %>% arrange(RMSE_log) %>% slice(1)
cat("Best model on TEST set:", best_model_row$Model, "\n")
```


```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

