---
title: "vole_project_"
author: "Elizabeth Binkina"
date: "4/15/2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)

# Read the data
df <- read_csv("processed_data.csv")
#View(df)

View(df)
```

```{r}
library(ggplot2)

ggplot(df, aes(x = Spp, fill = Spp)) +
  geom_bar() +
  facet_wrap(~ Site) +
  labs(title = "Histogram of MIPE and MIOC Counts by Site",
       x = "Species",
       y = "Frequency") +
  theme_minimal()
```

```{r}
library(ggplot2)

ggplot(df, aes(x = as.factor(Age), y = Actual_mass, fill = Spp)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, position = position_dodge(0.9)) +
  labs(title = "Distribution of Actual Mass by Age and Species",
       x = "Age",
       y = "Actual Mass") +
  theme_minimal() +
  scale_fill_manual(values = c("#c293cf", "#00a437"))
```


```{r}
library(RColorBrewer)

brewer.pal(3, "PRGn")
ggplot(df, aes(x = Site, y = Actual_mass, fill = Spp)) +
  geom_boxplot() +
  labs(title = "Distribution of Actual Mass by Site and Species",
       x = "Site",
       y = "Actual Mass") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("#c293cf", "#00a437"))  # PRGn purple and green

```

```{r}
df_juvenile <- subset(df, Age == "J")
df_adult <- subset(df, Age == "A")

df_juvenile
```



ANOVA SUMMARY






```{r}
model_juvenile <- aov(Actual_mass ~ Season * Spp * Y_coordinate , data = df_juvenile)
summary(model_juvenile)
```

```{r}
model_adult <- aov(Actual_mass ~ Season * Spp * Y_coordinate, data = df_adult)
summary(model_adult)

```

```{r}
par(mfrow = c(2, 2))
plot(model_juvenile)

# Adult
plot(model_adult)
```

```{r}
shapiro.test(residuals(model_juvenile))
shapiro.test(residuals(model_adult))

```

```{r}
kruskal.test(Actual_mass ~ Season, data = df_adult)
```

Model 1: Juveniles
ANOVA Results:
No significant effects of Season, Species (Spp), or their interaction on Actual_mass.

p-values for all terms > 0.2

Assumptions:
Shapiro-Wilk: W = 0.9833, p = 0.3178 → residuals are normally distributed 

Diagnostic Plots: Residuals appear homoscedastic and well-behaved.

Conclusion:

Juvenile mass is relatively stable across seasons and species.

No evidence of seasonal or species-related variation in juvenile body mass.

Juveniles might not be fully expressing seasonal or inter-species ecological adaptations that affect mass in adults.




Model 2: Adults
ANOVA Results:
Season is highly significant (F = 14.43, p < 0.001) 

Spp (Species) and interaction terms are not significant.

Assumptions:
Shapiro-Wilk: W = 0.99122, p = 0.0444 → slight deviation from normality 

Diagnostic Plots:

Slight tail skew in Q-Q plot, but generally acceptable.

No alarming patterns in residuals vs. fitted or scale-location.

Conclusion:
Adult body mass significantly varies across seasons.

This suggests strong seasonal physiological or behavioral changes??? --> consult with bio people
Species differences are not significant, indicating:

Either minimal size differences between species, or

That within-season variability swamps inter-species variation.

The interaction is also not significant → all species seem to follow a similar seasonal trend in mass.




```{r}
model_juvenile <- aov(Y_coordinate ~ Season * Spp * Actual_mass , data = df_juvenile)
summary(model_juvenile)
```

```{r}
model_adult <- aov(Y_coordinate ~ Season * Spp * Actual_mass , data = df_adult)
summary(model_adult)

```



```{r}
par(mfrow = c(2, 2))
plot(model_juvenile)

# Adult
plot(model_adult)
```

```{r}
df_juvenile$Season <- as.factor(df_juvenile$Season)
df_juvenile$Spp <- as.factor(df_juvenile$Spp)
model_juvenile_lm <- lm(Y_coordinate ~ Season * Spp * Actual_mass, data = df_juvenile)
summary(model_juvenile_lm)
```

```{r}
plot(model_juvenile_lm)
```

```{r}

df_adult$Season <- as.factor(df_adult$Season)
df_adult$Spp <- as.factor(df_adult$Spp)
model_adult_lm <- lm(Y_coordinate ~ Season * Spp * Actual_mass, data = df_juvenile)
summary(model_adult_lm)

```

```{r}

model_juv <- aov(Actual_mass ~  Spp * Y_coordinate, data = df_juvenile)
summary(model_juv)
```

```{r}
model_adult <- aov(Actual_mass ~ Spp * Y_coordinate , data = df_adult)
summary(model_adult)
```

```{r}
ggplot(df_juvenile, aes(x = Actual_mass, y = Y_coordinate, color = Spp)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, aes(linetype = Spp)) +
  facet_wrap(~ Season) +
  labs(
    title = "Interaction of Species, Season, and Mass on Y_coordinate (Juveniles)",
    x = "Actual Mass",
    y = "Y Coordinate (Latitude)",
    color = "Species",
    linetype = "Species"
  ) +
  theme_minimal()

```

```{r}

```

```{r}

```

```{r}

```


```{r}

```


```{r}

```






